---
title:  Regular expressions  
teaching: 15
exercises: 20
objectives:
- "Describe the difference between a text string and a regex pattern"
- "Use helper function `str_detect()` to find text"
- "Find an exact text match in a vector"
keypoints:
- "Regular expressions are the way we can describe text patterns to a computer"
- "Patterns can match either exact text or text that meets certain criteria"
- "The `str_detect()` function will search for a regex pattern within a text vector"
- "`.` is a wildcard pattern"
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("02-")
library(tidyverse)
```

## Regular expressions

Regular expressions (regex for short) are the 'language' used to specify patterns that can be used 
to search through text. The `stringr` package (part of the tidyverse) provides a consistent method
for working with regex patterns in R.

Regular expressions make use of symbols with special meaning to allow the freedom to describe
a pattern that is as specific or as ambiguous as required. Some uses for regular expressions can even
involve complex find and replace operations where the replacement text depends on the text pattern
that was matched.

> ## Other approaches
>
> > Some people, when confronted with a problem, think "I know, I'll use regular expressions." Now
> > they have two problems.
>
> Regular expressions are a very powerful tool, but can at times be complex to work with and difficult
> to read and understand in the future. There may be alternative approaches that can also achieve 
> the same result. Think about alternatives as you work through these challenges and consider whether
> regular expressions are the best option.
{: .callout}

## Vector refresher

When working with text data in R, it will probably be stored as a vector. Vectors in R are a collection
of values, that must all be of the same data type (see the 
[vector notes](https://csiro-data-school.github.io/r/16-Additional-content---Data-types/index.html#vectors) 
for a more detailed explanation). 

> ## Fruit vectors
> The `stringr` package has an inbuilt text vector for practicing text manipulation called `fruit`.
> Load the `stringr` package and explore the `fruit` vector to answer the following:
> 
> 1. How many elements are in the `fruit` vector?
> 2. What is the value of the first element?
> 3. What is the value of the last element?
{: .challenge}

## Finding fruit

At their simplest, regular expressions look to match the provided pattern exactly. This works exactly
the same as using the "Find" command in other programs. The `str_detect()` function can be used to
search for a pattern within some text. The first argument is the vector of text to be searched, and
the second argument is the pattern to search for (as a text string). The output of `str_detect()` 
is a `TRUE` or `FALSE` for each element of the text vector indicating whether the pattern was found.

For example, suppose we wanted to find the word 'apricot' in the `fruit` vector:

```{r finding-apricot}
str_detect(fruit, pattern = "apricot")
```

Since only the second element of `fruit` is `apricot`, only the second element of the output of 
`str_detect()` is `TRUE`.

> ## Understanding `str_detect()` 
> Run the following five commands. For each command, how many `TRUE`s are in the output? Can you
> explain why you get that result?
>
> 1. `str_detect(fruit, "apple")`
> 2. `str_detect(fruit, "rine")`
> 3. `str_detect(fruit, "Banana")`
> 4. `str_detect(fruit, "berry")`
> 5. `str_detect(fruit, " berry")`
>
{: .challenge}

> ## Working with `str_detect()`
> Write a `str_detect()` pattern that will find the following fruits:
> 
> 1. `bell pepper` and `chilli pepper`.
> 2. `clementine` and `cloudberry`.
> 3. `kiwi fruit`, `star fruit`, and `ugli fruit` (but not `breadfruit`, `dragonfruit`, etc.).
{: .challenge}

## Regular expression syntax

The real power in regular expressions comes from the ability to specify patterns that describe the
type of text to match, rather than trying to match text exactly. This allows you to match text that
has a common structure, but might have very different text.

> ## Specific currants
> In words, how could we identify 'blackcurrant' and 'redcurrant' from the list of fruits, 
> while ignoring the plain 'currant'? 
{: .challenge}

## Wildcards

Regular expression patterns have a number of special symbols that are used to describe the structure
of the text you are trying to match. The `.` character is a 'wildcard' symbol, and will match on *any*
text character, letter, number, whitespace, or symbol.

```{r using-wildcards}
# Matches blackcurrant, redcurrant, and currant
str_detect(fruit, "currant")

# Only matches blackcurrant and redcurrant
str_detect(fruit, ".currant")
```

> ## Using wildcards
> How many fruits have the letter `e` two characters after the letter `a`?
{: .challenge}

This wildcard symbol can be useful when you know the overall structure of the text you are trying to 
match, but the exact text is not important.

> ## What value?
> What are the potential benefits of an approach like this? Can you imagine any problems you might
> encounter?
> > ## Solution
> > Reproducible, repeatable, extensible, recorded and clear. 
> > Complex, truly extensible, truly clear/readable?? 
> {: .solution}
{: .discussion}

## Text in data frames
Like `str_detect()`, the regular expression functions we will be covering work on vectors of text 
data. Usually however, we have data stored in data frames. Within a data frame, each column is stored
as a separate vector. So we can use these regular expression functions within any function
that works on a column of a data frame (such as `mutate()`)

For example, if we put our `fruit` data into a data frame:

```{r make-a-fruit-df}
framed_fruit <- tibble(fruit_name = fruit)
```

we can create a new column with the results of `str_detect()` by using the name of a column with text
data as the first argument.

```{r mutate-fruit}
framed_fruit %>% 
  mutate(has_apple = str_detect(fruit_name, "apple"))
```

Since `str_detect()` provides a `TRUE` or `FALSE` result, it can even be used in a `filter()` 
function():

```{r filtering-fruit}
framed_fruit %>% 
  filter(str_detect(fruit_name, "apple"))
```


> ## Fruit frames
> Make a data frame containing the fruit data (as above).
> 
> Using the regex patterns for currants above, filter the data frame to the rows containing
> 
> 1. All currants (`blackcurrant`, `redcurrant`, and `currant`)
> 2. Just the `blackcurrant` and `redcurrant`
{: .challenge}
