---
title: Regular expression patterns
teaching: 30
exercises: 40
objectives:
- "Understand the basic features of regular expression patterns"
- "Use regex patterns to find specific text strings"
keypoints:
- "`|` will match the pattern on either side of the symbol"
- "To match any of a range of characters, include them within square brackets: `[]`"
- "`^` matches the start of a line, and `$` matches the end of a line"
- "Repeated matchings can be specified within `{}`, or using the `?`,`*` and `+` shortcuts"
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("03-")

library(tidyverse)
```

Regular expressions have a series of additional special symbols (like `.`) that allow you to more
specifically describe the text pattern you are searching for. These patterns can be as complex as required to identify the correct text in your data. 

To demonstrate the function of these symbols, we will work with the data on bird counts that was
downloaded earlier.

```{r reading-birds, message=FALSE}
birds <- read_csv("data/bird_counts.csv")

birds
```

## Alternative matches

The `|` character can be used as an 'OR' symbol, allowing matches from multiple options. For example,
the regex pattern `Rosella|Parrot`, would match at both the `Crimson Rosella` and `King-Parrot` values
in the `species` column.

```{r alternates}
# Alternative matching with |
birds %>% 
  mutate(is_colourful = str_detect(species, "Rosella|Parrot"))

# Can have multiple alternatives
birds %>% 
  mutate(not_cockatoo = str_detect(species, "Rosella|Parrot|Magpie"))
```

Any number of options can be specified by adding `|` symbols. But at least one *entire* option must
be found in the searched text for the regex pattern to be considered a match. 

For a single character match, options can be specified as a list within square brackets --`[]`. This
pattern will be considered a match if *any* of the elements of the list are found.

In addition to directly specifying the options (eg. `[abcde]`), they can also be provided as a range
of values (eg. `[a-e]`). Note that text matching is case sensitive, so `[a-z]` will match any lowercase
letter, but `[a-zA-Z]` will match any letter.

```{r square-brackets}
# Locations with an 'l' followed by a vowel
birds %>% 
  mutate(l_vowel = str_detect(location, "l[aeiou]"))

# Bird counts in double digits
birds %>% 
  mutate(double_digits = str_detect(count, "[0-9][0-9]"))
```

It is also possible to *not* match a list of characters by starting the list with a `^`.

```{r list-negation}
# Bird counts with non-numeric values
birds %>% 
  mutate(non_numeric = str_detect(count, "[^0-9]"))
```

> ## Alternative birds
> What is a regular expression pattern that would match:
> 
> 1. Text containing either `Magpie` or `cockatoo`
> 2. Text with either an `i`, `o` or `u` character
> 3. Any uppercase text character 
> 4. Any character *except* for `3`, `4`, or `5`.
>
> Using the bird count data, apply the first regex pattern to the `species` column using `mutate()`
> and `str_detect()`. 
>
> Then apply the second regex pattern to the `location` column, the third regex pattern to the 
> `count` column and the fourth pattern to the `date` column.
{: .challenge}

## Grouping and anchoring

There are a number of ways to control how these patterns are interpreted. 

`()` groups characters together so that the pattern inside is considered separately. This has many 
uses, for example to limit what alternatives are considered.

The regex pattern `Lake|BG` for example, will match either the text `Lake`, or the text `BG`. We can change the matching patterns by adding some grouping brackets. `L(ake|BG)` will match text that has
the letter `L`, then either `ake`, or `BG`. So it will match either the text `Lake`, or the text `LBG`.

```{r grouping}
birds %>% 
  mutate(lake = str_detect(location, "L(ake|BG)"))

birds %>% 
  mutate(gang_gangs = str_detect(species, "Gang(-| )gang"))
```

Using these groupings will become a lot more important later, as the text that is matched within the
brackets is captured and can be reused.

> ## More groupings
> Using `Gang( |-)gang` as a regex pattern works for rows two and eight, but fails to match the 
> fifth row of the birds data (with "Gang Gang cockatoo"). Edit this pattern to detect all 
> three rows.
{: .challenge}

The patterns can also be 'anchored' to specific parts of the text string being searched. This allows
us to limit where in the text we are searching for a match.

`^` is a special character meaning 'the start of a line'. (Different to how `^` is used within `[]`)
`$` is the opposite -- 'the end of a line'.

```{r anchoring}
# Match anywhere
birds %>% 
  mutate(ainslie = str_detect(location, "Ainslie"))

# Only match at start
birds %>% 
  mutate(starts_ainslie = str_detect(location, "^Ainslie"))

# Match anywhere
birds %>% 
  mutate(any_o = str_detect(species, "o"))

# Only match at end
birds %>% 
  mutate(ends_o = str_detect(species, "o$"))
```

> ## Getting more specific
> Using groupings with `()`, create a regex pattern that will match all the mountains in the 
> `location` column using `str_detect()`.
> 
> Modify your pattern above with an anchoring to exclude 'Black Mtn' from the matched values.
{: .challenge}

## Quantifying matches

As well as describing the text patterns to match, we can also specify how many times a pattern may
be repeated using `{}`.

* `{n}` matched exactly *n* times
* `{n,m}` matched between *n* and *m* times
* `{n, }` matched at least *n* times
* `{,m}` matched up to *m* times

```{r quant-exact}
birds %>% 
  mutate(double_o = str_detect(species, "o{2}"))
```

> ## What's being matched?
> As we move to more complicated examples it will be harder to identify problems with our patterns
> because `str_detect()` only returns a `TRUE` or `FALSE` value. Instead we will start to use 
> `str_extract()` which returns the actual text that is matched by the regex pattern.
{:.callout }

Be aware that this quantification repeats the *pattern*, not the matched text.

```{r quant-patterns}
# Does this match on any doubled letter?
birds %>% 
  mutate(any_doubled = str_detect(species, "[a-z]{2}"))

# No, it's the same as using [a-z][a-z]
birds %>% 
  mutate(any_doubled = str_extract(species, "[a-z]{2}"))
```

These apply to the regex pattern element immediately *before* the quantification pattern, which may
be a single character, bracked expression, or a grouping.

```{r quant-one-pattern}
birds %>% 
    mutate(four_to_six_letters = str_extract(location, "^[A-Z][a-z]{3,5} "))
```

Some of these quantifications are so commonly used that they have shortcut symbols:

* `?` matches 0-1 times -- so an *optional* pattern (same as `{0,1}`)
* `*` matches *any* number of times and may be absent (same as `{0,}`)
* `+` matches at least once (same as `{1,}`)

```{r quant-optional}
# Just like "Mt|Mtn"
birds %>% 
  mutate(short_mtns = str_extract(location, "Mtn?"))

# Can have multiple quantifiers
birds %>% 
  mutate(all_mtns = str_extract(location, "M(oun)?tn?"))
```

Quantified patterns will match as many times as possible up to their limit. This is called *greedy*
matching.

For example, the pattern `.+` (any character, one or more times) will match the entire text of a line.

```{r quant-greedy}
birds %>% 
  mutate(all = str_extract(species, ".+"))
```

Grouping patterns with `()` will often be necessary to make sure you are matching the correct text.

```{r quant-shortcuts}
# Doesn't match the full name
birds %>% 
  mutate(gang_gangs = str_extract(species, ".ang +"))

# But this one does
birds %>% 
  mutate(gang_gangs = str_extract(species, "(.ang )+"))
```

> ## Don't repeat yourself
> Using the birds data, write a regex pattern using `str_detect()` or `str_extract()` that will 
> match values in the `date` column that start with *two* digits, followed by a separator (`/` or `-`).
> 
> Can you modify this pattern so that it matches dates with two digits for the day, month, and year?
{: .challenge}
