---
title: Replacing text
teaching: 15
exercises: 30
questions: 
- "
- "
objectives:
- "Learn how to replace a regex pattern with specific text"
- "Explain how to replace multiple occurances of a pattern"
- "Use matched text in the replacement pattern"
keypoints:
- "`str_replace()` will find a regex pattern and replace it with the provided text"
- "If the regex pattern occurs multiple times in a string, `str_replace_all()` will replace all of them"
- "Backreferences can be used to include part of the matched pattern in the replacement text"
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("05-")

library(tidyverse)

birds <- read_csv("data/bird_counts.csv")
```

## Find and replace

Along with matching patterns or extracting text we can use regular expressions to perform find and
replace operations. By combining regex patterns and backreferences it is possible to build quite
complex substitutions should you need them.

To swap a matched regex pattern with some replacement text, we use the `str_replace()` function. This function requires three arguments, the text to search, what to find, and what to replace it with.

For example, say that you know the uninformative date in the birds data was actually recorded on 
19/1/20.

```{r simple-replace}
birds %>% 
  mutate(fixed_date = str_replace(date, "Last Sunday", "19/1/20"))
```

The text to find can use any of the regex patterns we have covered. It will be replaced in each row
where a match is found.

```{r replace-regex}
# Matches all rows
birds %>% 
  mutate(first_word = str_replace(location, "^\\w+", "First"))

# Matches most rows
birds %>% 
  mutate(first_word = str_replace(location, "^[A-Z][a-z]+", "First"))

```

> ## Replace once
> Using `str_replace()` and the birds data:
>
> 1. Replace the "Eight" in the `count` column with the character "8"
> 2. Replace all hyphens ("-") in the `species` column with spaces (" ")
> 3. Replace the digits for the year in the `date` column with the text "Year"
>
> > ## Solution
> > ``` {r replacing}
> > # Replace the "Eight" in the count column with the character "8"
> > birds %>% 
> >     mutate(replaced_data = str_replace(count, "Eight", "8"))
> > 
> > # Replace all hyphens in the species column with spaces
> > birds %>% 
> >     mutate(replaced_data = str_replace(species, "-", " "))
> > 
> > # Replace the digits for the year in the date column with the text "Year"
> > birds %>% 
> >     mutate(replaced_data = str_replace(date, "\\d+$", "Year"))
> > ```
> {: .solution}
{: .challenge}

## Replace more than once
When more than one match can be found in a text string, `str_replace()` will only replace the first
match.

```{r replace-once}
birds %>% 
  mutate(no_caps = str_replace(location, "[A-Z]", "_"))
```

If we want to replace all possible matches in our text, we need to use `str_replace_all()`

```{r}
birds %>% 
  mutate(no_caps = str_replace_all(location, "[A-Z]", "_"))
```

## Find and remove
You _could_ delete matched text by replacing it with an 'empty' string (`""`) using `str_replace()`.
However this is such a common task that there is a shortcut function in `str_remove()` (and
`str_remove_all()`).

```{r replace-delete}
birds %>% 
  mutate(devowel = str_remove_all(species, "[aeiou]"))
```

> ## Replace more
> Write a regex function that will:
>
> 1. Replace _each_ word in the `location` column with the text "word"
> 2. Replace any word seven letters or longer in the `species` column with the text "long"
> 3. Delete any non-numeric characters from the `count` column
> 4. Create an `abbreviation` column containing just the uppercase letters from the `location` column
>
> > ## Solution
> > ``` {r replacing-more}
> > # Replace each word in the location column with the text "word"
> > birds %>% 
> >     mutate(replaced_data = str_replace_all(location, "\\w+", "word"))
> > 
> > # Replace any word seven letters or longer in the species column with the text "long"
> > birds %>% 
> >     mutate(replaced_data = str_replace_all(species, "\\w{7,}", "long"))
> > 
> > # Delete any non-numeric characters from the count column
> > birds %>% 
> >     mutate(replaced_data = str_remove_all(count, "\\D"))
> >
> > # Create an abbreviation column containing just the uppercase letters from the location column
> > birds %>% 
> >     mutate(abbreviation = str_remove_all(location, "[^A-Z]"))
> > ```
> {: .solution}
{: .challenge}

## Use match in replacement
Just like in our regex patterns, we can use backreferences (`\1`, `\2`, etc.) to make use of 
'captured' text in our replacements.

For example, the regex pattern `(\w+)` will match and capture a complete word (`+`: one or more, 
`\w`: word characters). If we use the replacement pattern `*\1*`, it will replace the match with an
asterisk (`*`), followed by the matched word (`\1`), followed by another asterisk (`*`).

```{r backreference-replace}
birds %>% 
  mutate(fancy = str_replace_all(species, "(\\w+)", "*\\1*"))
```

As before, multiple matches can be referenced with increasing numbers.

```{r capture-more}
birds %>% 
  mutate(first_last = str_replace(species, "^(\\w).+(\\w)$", "First: \\1, Last: \\2"))
```

Note that the entire matched pattern gets replaced. In the example above, because we matched the whole
line with `.+` the entire line is replaced. If we wanted to keep that part of the text we would need
to capture it as well with grouping brackets.

```{r capture-all}
birds %>% 
  mutate(first_last = str_replace(species, "^(\\w)(.+)(\\w)$", "First: \\1, Last: \\3, Rest: \\2"))
```

> ## Complex replacements
> Using backreferences:
>
> 1. Double each vowel in the `location` column
> 2. Convert all dates in the `date` column to have four digit year values
> 3. Replace the middle letters of each word in the `species` column with two underscores. For example,
> `Crimson Rosella` should become `C__n R__a`
> 4. Print the first and last *word* of each `species` using the template "First: <word>, Last: <word>"
> (Tip: if you are having trouble, try including the boundary pattern -- `\b`)
>
> > ## Solution
> > ``` {r replacing-backreferences}
> > # Double each vowel in the location column
> > birds %>% 
> >     mutate(replaced_data = str_replace_all(location, "([aeiou])", "\\1\\1"))
> > 
> > # Convert all dates in the date column to have four digit year values
> > birds %>% 
> >     mutate(replaced_data = str_replace_all(date, "\\b(\\d{2})$", "20\\1"))
> > 
> > # Replace the middle letters of each word in the species column with two underscores
> > birds %>% 
> >     mutate(replaced_data = str_replace_all(species, "(\\w)\\w+(\\w)", "\\1__\\2"))
> >
> > # Print the first and last *word* of each location using the template "First: <word>, Last: <word>"
> > birds %>% 
> >     mutate(replaced_data = str_replace_all(species, "^(\\w+).+\\b(\\w+)$", "First: \\1, Last: \\2"))
> > ```
> {: .solution}
{: .challenge}

## Constructing a substitution

There are a few tips about the process that can help when trying to create a regex substution
pattern. We will demonstrate it with an example of wanting to change the date column from "day/month/year"
to "year:month:day (old: day/month/year)"

1. Start by copying a real example of a whole string to match as your pattern

**Pattern**: `21/12/19`

2. Add any escape backslashes necessary of your text contains regex special symbols (not needed in this
example)

**Pattern**: `21/12/19`

3. Circle the parts of the string you’d like to separately retain, with round brackets.

**Pattern**: `((21)/(12)/(19))`

4. Write out your replacement pattern, using backreferences to what you circled.

**Pattern**: `((21)/(12)/(19))`, **Replacement**: `\4:\3:\2 (old: \1)`

5. At this stage, the substitution should work, but only for the specific real example string that you’ve started with. Test it to make sure.

```{r test-sub}
birds %>% 
  mutate(test_replace = str_replace(date, "((21)/(12)/(19))", "\\4:\\3:\\2 (old: \\1)"))
```

6. Finally, start abstracting your search pattern, replacing parts of your example string with wild-cards or character-classes as needed, to strike the balance between specificity and ambiguity required to match all that you want and not all that you don’t want.

You may need to test your abstracted pattern often to make sure you are matching the right text. There
will often be multiple regex patterns that match the 'correct' text. Your future self will thank you
if you can come up with a pattern that is easier to read.

Using digit/non-digit distinction as a 'simple' solution:
**Pattern**: `((\d+)\D(\d+)\D(\d+))`, **Replacement**: `\4:\3:\2 (old: \1)`

```{r correct-match}
birds %>% 
  mutate(changed_date = str_replace(date, "((\\d+)\\D(\\d+)\\D(\\d+))", "\\4:\\3:\\2 (old: \\1)"))
```

Can get more specific, for example adding error checking using number of digits. But you trade off 
accuracy of matching for complexity, so consider how complex a pattern needs to be to match the sort
of data you are likely to receive.

**Pattern**: `((\d{1,2})\D(\d{1,2})\D(\d{2}|\d{4}))`, **Replacement**: `\4:\3:\2 (old: \1)`

> ## Putting it together
> Create a cleaned version of the birds data called `clean_birds`. You will need to use multiple steps
> and the regex functions `str_extract()`, `str_replace()`, and `str_remove()` (or their `XXXX_all()`
> variants).
>
> A cleaned data frame should have:
>
> * Consistent naming for the `location` and `species` column so that each location/species has only
> a single way of being recorded
> * A consistent format for the `date`column 
> * Additional information from the `count` column put in a new column called `notes`
> * Only numeric values in the `count` column
> * Any other problems you noticed in the 
> [first challenge](../01-messy-data/index.html#bird-sightings-in-canberra) that have not otherwise
> been fixed
{: .challenge}